Long-term trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013

### 2017-06-29 Manuscript/Manuscript_COD_Trends_BMJOpen.Rmd
changed ref order

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in 39 US states and the District of Columbia between 1969 and 2013.^28^ We excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming. These states contained less than 0.4% of the black population in the United States in 2010.^5^ The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.^29^ The population data were derived by the NCHS using intercensal bridged-race population estimates. We used data on the number of deaths and the population size according to race (black or white), gender (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**Table S1**). We did not consider Hispanic ethnicity as part of our racial definition, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census.^6^

@@

The data structure produced two statistical challenges that were addressed by our modeling strategy. Firstly, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9,^30^ that we needed to impute. Secondly, several states had small black populations which produced sparse counts in some state-year-race-sex-age-cause strata. In these cases, the estimation of the mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying mortality rates, we used a method that produces smoother age-specific mortality trends.

@@

To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total deaths in each age group, and standard life table methods were applied.^31,32^ Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and gender.

To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method.^33^ This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher cause-specific mortality rates among blacks. For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher rate of mortality due to cardiovascular disease among blacks compared to whites. If the decomposition produces a negative estimate this implies that whites have a higher rate of mortality for injuries compared to blacks. Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 40 state-level trends and their credible intervals.

We also calculated state-level time trends in age-standardized cause-specific mortality rates separately for each combination of race and gender, to aid the interpretation of the decomposition results. We used the US standard population from the year 2000.^34^

@@

We examined state-level trends in six major cause-of-death contributions to the black-white life expectancy gap over a 45-year period. While we found that many states conformed to the national pattern of cause-of-death specific contributions, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation. Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West, which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20th Century.^9^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi. On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^8^ On the other hand, greater racial equality can be achieved by falling black mortality rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.^10^

Consistent with what has been reported at the national level,^11^ we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time. Injuries (especially homicide) continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South. This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic.^12^ Communicable disease continues to contribute to the gap in most states, especially in DC, New Jersey, and Florida. In our previous work, approximately half of the contribution of communicable disease to the gap was found to be due to HIV in 2008 nationally,^11^ suggesting that these states may increase equality through interventions targeted at reducing mortality from HIV. Cancer is also responsible for large contributions to the life expectancy gap in some states (Illinnois, Michigan, and Wisconsin) but not in others (New York, Massachusetts). Overall, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women. However, both states began with only a minimal contribution of CVD and cancer to the mortality gap. The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contributions of more than two years. Thus, the reason for New York’s and Massachusetts’ low CVD contributions does not appear to be the effect of recent changes, but highlights a long-standing pattern of greater equality and points to the importance of viewing the most recent estimates in light of historical patterns.

@@

We used data over 45 years, across two racial groups, and for all ages. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause of death were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time. Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.^19,20^ The Arriaga decomposition method accounts for the fact that deaths occurring in one cause stratum are reflected in the absence of that death in another stratum. In contrast to some recent work focused on county-level inequalities,^7,8^ states have some substantial advantages as analysis units; They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities, and so have divergent social agendas that are reflected in varying patterns of mortality by race.

Our analysis has three main limitations. Firstly, we did not account for Hispanic ethnicity, since ethnicity was not measured until 1978, and even then it was poorly measured on the death certificate in many states for many years and our previous sensitivity analysis including Hispanics did not yield a substantial difference in the results.^6^, 4 Had Hispanics been excluded from this study, the life expectancy estimates for both blacks and whites are expected to be lower, since Hispanics are known to have higher life expectancies than blacks and whites.^22^ Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states. US immigrants have been shown to have life expectancies at birth that were approximately 5 years longer in 2010.^24^ Thus, the black-white gap in a state may decrease as more immigrants identify as black. Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death. Migration is particularly important for the District of Columbia, because of its small geographical bounds. Affluent black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC. Both changes would serve to increase the racial gap in DC.^25^ Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes. For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time. Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.^26,27^

@@

The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time. States pursue varying approaches to an array of relevant programs and laws including tobacco control,^13^ Medicare coverage,^14^ hand-gun violence,^15^ and nutrition.^16^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.^17^ The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others. Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations.

@@

5. U.S. Census Bureau: United States Census 2010 Interactive Population Map [Internet]. [cited 2017 May 16]. Available from: https://www.census.gov/2010census/popmap/.

6. Arias E, Schauman WS, Eschbach K, Sorlie PD, Backlund E (2008) The validity of race and Hispanic origin reporting on death certificates in the United States. National Center for Health Statistics. Vital Health Stat 2(148). Available from: https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf

7. Arora A, Spatz E, Herrin J, Riley C, Roy B, Kell K, Coberley C, Rula E, Krumholz HM (2016) Population Well-Being Measures Help Explain Geographic Disparities In Life Expectancy At The County Level. *Health Aff* 35(11):2075-82.

8. Dwyer-Lindgren L, Bertozzi-Villa A, Stubbs RW, Morozoff C, Kutz MJ, Huynh C, Barber RM, et al (2016) US County-Level Trends in Mortality Rates for Major Causes of Death, 1980-2014. *JAMA* 316(22):2385–2401.

9. Lemann, N (1991) The Promised Land: The Great Black Migration and How It Changed America. New York: Alfred A. Knopf, ISBN 0-394-56004-3.

10. Case A, Deaton A (2017) Mortality and Morbidity in the 21st Century. In: Eberly J and Stock J, Editors. Brookings Papers on Economic Activity, March 23-24 2017.

11. Harper S, Rushani D, and Kaufman JS (2012) Trends in the Black-White Life Expectancy Gap, 2003-2008. *JAMA* 307(21):2257–9.

12. Rudd RA, Seth P, David F, Scholl L (2016) Increases in Drug and Opioid-Involved Overdose Deaths — United States, 2010–2015. *MMWR Morb Mortal Wkly Rep* 65:1445–1452.

13. Cotti C, Nesson E, Tefft N (2016) The Effects of Tobacco Control Policies on Tobacco Products, Tar, and Nicotine Purchases among Adults: Evidence from Household Panel Data. *American Economic Journal: Economic Policy* 8(4):103-23.

14. Grogan CM, Park SE (2017) The Racial Divide in State Medicaid Expansions. *J Health Polit Policy Law* 42(3):539-72.

15. Humphreys DK, Gasparrini A, Wiebe DJ (2017) Evaluating the Impact of Florida’s “Stand Your Ground” Self-defense Law on Homicide and Suicide by Firearm: An Interrupted Time Series Study. *JAMA Intern Med* 177(1):44-50.

16. Taber DR, Stevens J, Evenson KR, Ward DS, Poole C, Maciejewski ML, Murray DM, Brownson RC (2011) State policies targeting junk food in schools: racial/ethnic differences in the effect of policy change on soda consumption. *Am J Public Health* 101(9):1769-75.

17. Adimora AA, Ramirez C, Schoenbach VJ, Cohen MS (2014) Policies and politics that promote HIV infection in the Southern United States. *AIDS*
28(10):1393.

18. Strumpf EC, Harper S, Kaufman JS (2017) Fixed Effects and Difference in Differences. In: Methods in Social Epidemiology. 2nd Ed, eds Oakes M, Kaufman JS(Jossey-Bass, San Francisco).

@@

21. LaVeist-Ramos TA, Galarraga J, Thorpe RJ, Bell CN, Austin CJ (2011) Are black Hispanics black or Hispanic? Exploring disparities at the intersection of race and ethnicity. *J Epidemiol Community Health* Mar 3:jech-2009.

@@

23. Brazil N (2017) Spatial Variation in the Hispanic Paradox: Mortality Rates in New and Established Hispanic US Destinations. *Population, Space and Place* 23:e1968.

24. Preston SH, Elo IT (2014) Anatomy of a municipal triumph: New York City's upsurge in life expectancy. *Population and development review* 40(1):1-29.

25. U.S. Census Bureau. Census Reporter Profile Page: American Community Survey 1-year estimates [Internet]. [Cited 16 May 2017]. Available from: https://censusreporter.org/profiles/16000US1150000-washington-dc/.

26. Kuehn BM (2014) CDC: major disparities in opioid prescribing among states: some states crack down on excess prescribing. *JAMA* 312(7):684-6.

27. Jones CM, Mack KA, Paulozzi LJ (2010) Pharmaceutical overdose deaths, United States, 2010. *JAMA* 309(7):657-9.

28. National Cancer Institute. Surveillance, Epidemiology, and End Results
Program: SEER*Stat software: release: version 8.3.2 [Internet]. Bethesda (MD): NCI; 2017 Apr 14 [cited 2016 August 20]. Available from: http://www.seer.cancer.gov/seerstat.

29. Centers for Disease Control and Prevention. National Center for Health Statistics: National Vital Statistics System [Internet]. [cited 2017 May 16]. Available from: https://www.cdc.gov/nchs/nvss/index.htm.

30. National Cancer Institute. Surveillance, Epidemiology, and End Results Program: US mortality data [Internet]. Bethesda (MD): NCI; [cited 2017 May 16]. Available from: http://www.seer.cancer.gov/mortality/

31. Preston SH, Heuveline P, and Guillot M (2001) Demography: Measuring and Modeling Population Processes. (Blackwell Publishing, Malden)

32. Selvin, S (2004) Statistical Analysis of Epidemiologic Data. 3rd ed. (Oxford University Press, Oxford)

33. Arriaga EE (1985) Measuring and explaining the change in life expectancies. *Demography* 21(1):83-96.

34. National Cancer Institute. Surveillance, Epidemiology, and End Results Program: Standard Populations - 19 Age Groups [Internet]. [cited 2017 April 25]. Available from: https://seer.cancer.gov/stdpopulations/stdpop.19ages.html.

>>>>

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in 39 US states and the District of Columbia between 1969 and 2013.^5^ We excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming. These states contained less than 0.4% of the black population in the United States in 2010.^6^ The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.^7^ The population data were derived by the NCHS using intercensal bridged-race population estimates. We used data on the number of deaths and the population size according to race (black or white), gender (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**Table S1**). We did not consider Hispanic ethnicity as part of our racial definition, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census.^8^

@@

The data structure produced two statistical challenges that were addressed by our modeling strategy. Firstly, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9,^9^ that we needed to impute. Secondly, several states had small black populations which produced sparse counts in some state-year-race-sex-age-cause strata. In these cases, the estimation of the mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying mortality rates, we used a method that produces smoother age-specific mortality trends.

@@

To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total deaths in each age group, and standard life table methods were applied.^10,11^ Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and gender.

To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method.^12^ This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher cause-specific mortality rates among blacks. For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher rate of mortality due to cardiovascular disease among blacks compared to whites. If the decomposition produces a negative estimate this implies that whites have a higher rate of mortality for injuries compared to blacks. Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 40 state-level trends and their credible intervals.

We also calculated state-level time trends in age-standardized cause-specific mortality rates separately for each combination of race and gender, to aid the interpretation of the decomposition results. We used the US standard population from the year 2000.^13^

@@

We examined state-level trends in six major cause-of-death contributions to the black-white life expectancy gap over a 45-year period. While we found that many states conformed to the national pattern of cause-of-death specific contributions, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation. Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West, which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20th Century.^14^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi. On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^15^ On the other hand, greater racial equality can be achieved by falling black mortality rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.^16^

Consistent with what has been reported at the national level,^17^ we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time. Injuries (especially homicide) continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South. This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic.^18^ Communicable disease continues to contribute to the gap in most states, especially in DC, New Jersey, and Florida. In our previous work, approximately half of the contribution of communicable disease to the gap was found to be due to HIV in 2008 nationally,^17^ suggesting that these states may increase equality through interventions targeted at reducing mortality from HIV. Cancer is also responsible for large contributions to the life expectancy gap in some states (Illinnois, Michigan, and Wisconsin) but not in others (New York, Massachusetts). Overall, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women. However, both states began with only a minimal contribution of CVD and cancer to the mortality gap. The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contributions of more than two years. Thus, the reason for New York’s and Massachusetts’ low CVD contributions does not appear to be the effect of recent changes, but highlights a long-standing pattern of greater equality and points to the importance of viewing the most recent estimates in light of historical patterns.

@@

We used data over 45 years, across two racial groups, and for all ages. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause of death were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time. Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.^19,20^ The Arriaga decomposition method accounts for the fact that deaths occurring in one cause stratum are reflected in the absence of that death in another stratum. In contrast to some recent work focused on county-level inequalities,^21,15^ states have some substantial advantages as analysis units; They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities, and so have divergent social agendas that are reflected in varying patterns of mortality by race.

Our analysis has three main limitations. Firstly, we did not account for Hispanic ethnicity, since ethnicity was not measured until 1978, and even then it was poorly measured on the death certificate in many states for many years and our previous sensitivity analysis including Hispanics did not yield a substantial difference in the results.^8^ Had Hispanics been excluded from this study, the life expectancy estimates for both blacks and whites are expected to be lower, since Hispanics are known to have higher life expectancies than blacks and whites.^22^ Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states. US immigrants have been shown to have life expectancies at birth that were approximately 5 years longer in 2010.^23^ Thus, the black-white gap in a state may decrease as more immigrants identify as black. Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death. Migration is particularly important for the District of Columbia, because of its small geographical bounds. Affluent black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC. Both changes would serve to increase the racial gap in DC.^24^ Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes. For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time. Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.^25,26^

@@

The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time. States pursue varying approaches to an array of relevant programs and laws including tobacco control,^27^ Medicare coverage,^28^ hand-gun violence,^29^ and nutrition.^30^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.^31^ The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others. Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations.^32^

@@

5. National Cancer Institute. Surveillance, Epidemiology, and End Results
Program: SEER*Stat software: release: version 8.3.2 [Internet]. Bethesda (MD): NCI; 2017 Apr 14 [cited 2016 August 20]. Available from: http://www.seer.cancer.gov/seerstat.

6. U.S. Census Bureau: United States Census 2010 Interactive Population Map [Internet]. [cited 2017 May 16]. Available from: https://www.census.gov/2010census/popmap/.

7. Centers for Disease Control and Prevention. National Center for Health Statistics: National Vital Statistics System [Internet]. [cited 2017 May 16]. Available from: https://www.cdc.gov/nchs/nvss/index.htm.

8. Arias E, Schauman WS, Eschbach K, Sorlie PD, Backlund E (2008) The validity of race and Hispanic origin reporting on death certificates in the United States. National Center for Health Statistics. Vital Health Stat 2(148). Available from: https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf

9. National Cancer Institute. Surveillance, Epidemiology, and End Results Program: US mortality data [Internet]. Bethesda (MD): NCI; [cited 2017 May 16]. Available from: http://www.seer.cancer.gov/mortality/

10. Preston SH, Heuveline P, and Guillot M (2001) Demography: Measuring and Modeling Population Processes. (Blackwell Publishing, Malden)

11. Selvin, S (2004) Statistical Analysis of Epidemiologic Data. 3rd ed. (Oxford University Press, Oxford)

12. Arriaga EE (1985) Measuring and explaining the change in life expectancies. *Demography* 21(1):83-96.

13. National Cancer Institute. Surveillance, Epidemiology, and End Results Program: Standard Populations - 19 Age Groups [Internet]. [cited 2017 April 25]. Available from: https://seer.cancer.gov/stdpopulations/stdpop.19ages.html.

14. Lemann, N (1991) The Promised Land: The Great Black Migration and How It Changed America. New York: Alfred A. Knopf, ISBN 0-394-56004-3.

15. Dwyer-Lindgren L, Bertozzi-Villa A, Stubbs RW, Morozoff C, Kutz MJ, Huynh C, Barber RM, et al (2016) US County-Level Trends in Mortality Rates for Major Causes of Death, 1980-2014. *JAMA* 316(22):2385–2401.

16. Case A, Deaton A (2017) Mortality and Morbidity in the 21st Century. In: Eberly J and Stock J, Editors. Brookings Papers on Economic Activity, March 23-24 2017.

17. Harper S, Rushani D, and Kaufman JS (2012) Trends in the Black-White Life Expectancy Gap, 2003-2008. *JAMA* 307(21):2257–9.

18. Rudd RA, Seth P, David F, Scholl L (2016) Increases in Drug and Opioid-Involved Overdose Deaths — United States, 2010–2015. *MMWR Morb Mortal Wkly Rep* 65:1445–1452.

@@

21. Arora A, Spatz E, Herrin J, Riley C, Roy B, Kell K, Coberley C, Rula E, Krumholz HM (2016) Population Well-Being Measures Help Explain Geographic Disparities In Life Expectancy At The County Level. *Health Aff* 35(11):2075-82.

@@

23. Preston SH, Elo IT (2014) Anatomy of a municipal triumph: New York City's upsurge in life expectancy. *Population and development review* 40(1):1-29.

24. U.S. Census Bureau. Census Reporter Profile Page: American Community Survey 1-year estimates [Internet]. [Cited 16 May 2017]. Available from: https://censusreporter.org/profiles/16000US1150000-washington-dc/.

25. Kuehn BM (2014) CDC: major disparities in opioid prescribing among states: some states crack down on excess prescribing. *JAMA* 312(7):684-6.

26. Jones CM, Mack KA, Paulozzi LJ (2010) Pharmaceutical overdose deaths, United States, 2010. *JAMA* 309(7):657-9.

27. Cotti C, Nesson E, Tefft N (2016) The Effects of Tobacco Control Policies on Tobacco Products, Tar, and Nicotine Purchases among Adults: Evidence from Household Panel Data. *American Economic Journal: Economic Policy* 8(4):103-23.

28. Grogan CM, Park SE (2017) The Racial Divide in State Medicaid Expansions. *J Health Polit Policy Law* 42(3):539-72.

29. Humphreys DK, Gasparrini A, Wiebe DJ (2017) Evaluating the Impact of Florida’s “Stand Your Ground” Self-defense Law on Homicide and Suicide by Firearm: An Interrupted Time Series Study. *JAMA Intern Med* 177(1):44-50.

30. Taber DR, Stevens J, Evenson KR, Ward DS, Poole C, Maciejewski ML, Murray DM, Brownson RC (2011) State policies targeting junk food in schools: racial/ethnic differences in the effect of policy change on soda consumption. *Am J Public Health* 101(9):1769-75.

31. Adimora AA, Ramirez C, Schoenbach VJ, Cohen MS (2014) Policies and politics that promote HIV infection in the Southern United States. *AIDS*
28(10):1393.

32. Strumpf EC, Harper S, Kaufman JS (2017) Fixed Effects and Difference in Differences. In: Methods in Social Epidemiology. 2nd Ed, eds Oakes M, Kaufman JS(Jossey-Bass, San Francisco).
